---
output:
  pdf_document: default
  html_document: default
---
#  Linear Regression

##### Josep Fortiana 2019-10-01

## 1. Simple linear regression: A simulated example

### Generate a simulated dataset for simple linear regression

```{r}
# More or less arbitrary parameters for the data generation

trueA<-1   
trueB<-5
trueSd<-15
sampleSize<-20 # n = sampleSize 
```

```{r}
# Uniformly spread x values in the interval [-n/2,n/2]
x<-(-(sampleSize-1)/2):((sampleSize-1)/2)
# y values of the form a + b*x + N(0,trueSd)
y<-trueA+trueB*x+rnorm(n=sampleSize,mean=0,sd=trueSd)
```

```{r}
# Plot data points plus the line used in the generation
# Note this line is NOT a regression line, which will be computed below
options(repr.plot.width=4, repr.plot.height=4)
plot(x,y,pch='+',cex=0.8,col="blue",main="Test data")
abline(trueA,trueB,lwd=2.5,col="cyan")
```

### Adjust a least squares linear regression model

```{r}
lm.1<-lm(y~x)
options(repr.plot.width=4, repr.plot.height=4)
plot(x,y,pch='+',cex=0.8,col="blue",main="Test data")
abline(trueA,trueB,lwd=2.5,col="cyan")
abline(lm.1,lwd=2.5,col="blue")
legend("topleft",c("Original line","","Regression line"), lwd=2.5,col=c("cyan","white","blue"),cex=0.6)
```

### Extract information from the fitted model

```{r}
print(lm.1)
summary(lm.1)
```

### Regression coefficients

```{r}
Coeff<-coefficients(lm.1)
Coeff
a.hat<-Coeff[1]
b.hat<-Coeff[2]
#a.hat<-as.numeric(lm.1$coefficients[1])
#b.hat<-as.numeric(lm.1$coefficients[2])
round(a.hat,3)
round(b.hat,3)
```

## 2. Relevant quantities in a regression

### 1. Total Sum of Squares

```{r}
# Total sum of squares
TotalSS<-sum(y^2)
round(TotalSS,3)
# Centered data
y0<-y-mean(y)   
# Centered total sum of squares
TotalSS0<-sum(y0^2)        
round(TotalSS0,3)
# Total number of degrees of freedom
n<-length(y)
Totaldf<-n
# Total number of degrees of freedom of the centered y
Totaldf0<-n-1
Totaldf
Totaldf0
```

### 2. Fitted values and Regression Sum of Squares

```{r}
# Fitted values
yhat<-fitted.values(lm.1)
#yhat<-as.numeric(lm.1$fitted.values) # Alternative syntax
# Centered fitted values
yhat0<-yhat-mean(yhat)
# Regression Sum of Squares
RegSS<-sum(yhat^2)
round(RegSS,3)
# Centered Regression Sum if Squares
RegSS0<-sum(yhat0^2)
round(RegSS0,3)
# Number of degrees of freedom of the regression
Regdf<-2
# Number of degrees of freedom of the regression (centered)
Regdf0<-1

Regdf
Regdf0
```

Check that `mean(y)` coincides with `mean(yhat)`

### 3. Regression residuals and Residual Sum of Squares

```{r}
# The regression residuals can be extracted as:
ytilde<-residuals(lm.1)
# ytilde<-as.numeric(lm.11$residuals) # Alternative syntax
# Also: 
# ytilde<-y-yhat 
# or
# ytilde<-y0-yhat0
#
# Note that, since both y and yhat have the same mean, regression residuals are centered.
ResSS<-sum(ytilde^2)
Resdf<-Totaldf0-Regdf0
round(ResSS,3)
Resdf
```

### 4. [Total Sum of Squares]  =  [Regression Sum of Squares]  +  [Residual Sum of Squares]

```{r}
# Both with the non-centered and with the centered version
round(TotalSS-(RegSS+ResSS),10)
round(TotalSS0-(RegSS0+ResSS),10)
```

### 5. Regression Coefficient of Determination (Multiple R-squared)

```{r}
# By definition, with the centered sums of squares
R2<-RegSS0/TotalSS0
round(R2,4)
```

### 6. Adjusted Coefficient of Determination (Adjusted for the number $p$ of predictors)

$$
    {\bar {R}}^{2}=1-(1-R^{2})\dfrac{n-1}{n-p-1}
$$

The Adjusted Coefficient of Determination is also equal to:

$$
    {\displaystyle {\bar {R}}^{2}={1-{SS_{\text{res}}/{\text{df}}_{e} \over SS_{\text{tot}0}/{\text{df}}_{t0}}}}
$$

```{r}
p<-1
R2adj.1<-1-(1-R2)*(n-1)/(n-p-1)
round(R2adj.1,4)
R2adj.2<-1-(ResSS/Resdf)/(TotalSS0/Totaldf0)
round(R2adj.1,4)
```

### 7. Mean squares and regression F statistic

```{r}
TotalMeanS0<-TotalSS0/Totaldf0
RegMeanS0<-RegSS0/Regdf0
ResMeanS<-ResSS/Resdf    # Remember that residuals are centered (hence there is no need of a "0" here)
F<-RegMeanS0/ResMeanS
# The p-value is the probability of obtaining F values larger than the observed one, assuming the null 
# hypothesis that there is no regression relationship is true.
p.val<-1-pf(F,df1=Regdf0,df2=Resdf)
round(TotalMeanS0,3)
round(RegMeanS0,3)
round(ResMeanS,3)
round(F,3)
round(p.val,10)
```

The quotient F is a measure of how the Regression Mean Squares exceeds the Residual Mean of Squares. 

When the model is a Gauss-Markov normal regression, this quantity follows a Fisher-Snedecor distributions with degrees of freedom Regdf0 and Resdf. The resulting $p$-value is computed assuming this is true. When it is larger than the standard significance level ($p$-value > 0.05) we conclude the regression model is non-significant.

### 8. The `anova()` function

Displays the Sums of Squares and Mean Squares, decomposed by each individual predictor and residuals contribution (here there is a single predictor `x`).

```{r}
lm.1.anova<-anova(lm.1)
lm.1.anova
# str(lm.1.anova)
```

### 9. Residual standard error (estimate $\hat{\sigma}$ of the residuals standard deviation)

```{r}
# Computing from the relevant quantities we have:
ResSE<-sqrt(ResMeanS)
round(ResSE,3)
# Alternatively, from the anova() function output:
sigma2.hat<-lm.1.anova$Sum[2]/lm.1.anova$Df[2]
sigma.hat<-sqrt(sigma2.hat)
round(sigma.hat,3)
```

### 10. Estimate of the matrix of variances and covariances of  $\hat{a}$ and $\hat{b}$

```{r}
V<-vcov(lm.1)
round(V,3)
# Null out-of-diagonal entries result from the peculiar way x was constructed for this example. 
# In general one would expect non-null entries.
```

### 11. Estimates of the standard deviations of the regression coefficients

```{r}
v<-diag(V)
sigma2.a<-as.numeric(v[1])
sigma2.b<-as.numeric(v[2])
sigma.a<-sqrt(sigma2.a)
sigma.b<-sqrt(sigma2.b)
round(sigma.a,4)
round(sigma.b,4)
```

### 12. Student's $t$ statistics

They are a standardized measure of how each of the regression coefficients, here $\hat{a}$ and $\hat{b}$, differ from zero.

When data are normal and the the regression model is Gauss-Màrkov these quantities follow a Student's distribution $t(n-p-1)$. 

For each coefficient, the $p$-value is, in principle, used to test the hypotheses:

$$
\left\{
\begin{array}{cl}
    H_{0}: \mskip15mu&\text{The coefficient is null,}\\
    H_{1}: \mskip15mu&\text{The coefficient is not zero,}
\end{array}\right.
$$

In practice, these $p$-valors are taken as a mere hint, a first step to a more precise assessment of the importance of each predictor.

```{r}
t.a<-as.numeric(a.hat/sigma.a)
t.b<-as.numeric(b.hat/sigma.b)
round(t.a,4)
round(t.b,4)
p.val.a<-2*pt(-abs(t.a),n-2)
p.val.b<-2*pt(-abs(t.b),n-2)
round(p.val.a,8)
round(p.val.b,12)
S<-summary(lm.1)
coefficients(S)
```

### 13. The regression matrix (or model matrix)

Now we see how the least squares regression is actually computed:

```{r}
X<-model.matrix(lm.1)
str(X)
X
```

### 14 . Regression coefficients (estimates of)

The regression coefficients vector $\hat{\beta}=\left(\begin{array}{c}\hat{a}\\\hat{b}\end{array}\right)$ is computed as:

$$
    \hat{\beta}=(X'\cdot X)^{-1}\cdot X' \cdot y
$$

```{r}
# Matrix to invert
Q<-t(X) %*% X
round(Q,3)
```

```{r}
# Compute regression coefficients:
Q1<-solve(Q)   # inverse matrix
beta.hat<-Q1 %*%t(X) %*% y
round(beta.hat,4)
```

### 15. The hat matrix and the fitted $\hat{y}$ values

The hat matrix for a regression with model matrix $X$ is:

$$
    H=X\cdot(X'\cdot X)^{-1}\cdot X'.
$$

It satisfies that:

$$
    H\cdot y  = X\cdot(X'\cdot X)^{-1}\cdot X'\cdot y= X\cdot(X'\cdot X)^{-1}\cdot X'\cdot y = X\cdot \hat{\beta} = \hat{y}.
$$

```{r}
# The hat matrix is the [n,n] matrix:
H<-X %*% solve(Q)%*%t(X)
# H is the operator that "puts a hat" on y, giving yhat. 
```

```{r}
# Check that indeed this yhat coincides with the one obtained above
yhat.1<-H %*% y
round(max(abs(yhat-yhat.1)),15)
```

Properties of $H$

```{r}
# The sum of diagonal entries in H (the trace of H) is equal to the rank of X
sum(diag(H))
# H is an idempotent matrix. The square of H is equal to H.
round(max(abs(H %*% H-H)),15)
#round(H,3)
```

### 16. Covariances of regression coefficients estimates

The matrix of variances and covariances of the coefficients vector:

$$
    \hat{\beta}=(X'\cdot X)^{-1}\cdot X' \cdot y
$$

is computed by:

$$
    \operatorname{Var}(\hat{\beta})=(X'\cdot X)^{-1}\cdot X'
       \cdot\operatorname{Var}(y)\cdot X\cdot(X'\cdot X)^{-1}
$$

If the Gauss-Markov condition holds, then $\operatorname{Var}(y)=\sigma^2\,I$ and the expression above is simplified:

$$
      \operatorname{Var}(\hat{\beta})=\sigma^2\,(X'\cdot X)^{-1}
$$

```{r}
# Compare this V1 with the one obtained above with the vcov() function
V1<-Q1*sigma2.hat
V1
```

## 3. The `Advertising` dataset

Download `Advertising.csv`, from the textbook web page [An Introduction to Statistical Learning with Applications in R (ISLR)](http://www-bcf.usc.edu/~gareth/ISL/data.html). This dataset is used both in lesson 3 in this course and in Chapter 3 of the textbook. In this section we follow the treatment of the `Advertising` dataset from his chapter.

Set the directory where you saved the dataset file as the RStudio working directory.

Read dataset.

Omit the first column, variable `X`, the index of each sample. Anyway it can be recovered by: `row.names(Advertising)`.

```{r}
Advertising<-read.csv("Advertising.csv")
Advertising<-Advertising[,-1]
```

### Pairwise simple linear regressions

Following ISLR, Chapter 3:

Compute simple linear regressions, one for each of the predictors, `TV`, `Radio`, `Newspapers`. 

Plot all three scatterplots, superimposing on them the regression line.

What can be said about goodness-of-fit of these models?

Which variable, of the possible predictors, `TV`, `Radio`, `Newspapers`, is a better predictor of `Sales`?

Compare the goodness of fit of these linear regressions with the k-NN regressions from last lesson.

```{r}
#
# Insert your code here
#
```

## 4. Multiple regression

```{r}
Advertising<-read.csv("Advertising.csv")
Advertising<-Advertising[,-1]
```

```{r}
str(Advertising)
```

```{r}
plot(Advertising)
```

```{r}
cor(Advertising)
round(cor(Advertising),2)
round(cor(Advertising),1)
```

Multiple linear regression of Sales on all three predictors

```{r}
lm.Advertising.01<-lm(Sales~TV+Radio+Newspaper,data=Advertising)
summary(lm.Advertising.01)
anova(lm.Advertising.01)
```

```{r}
# Equivalent, alternative, notation
lm.Advertising.02<-lm(Sales~.,data=Advertising)
summary(lm.Advertising.02)
anova(lm.Advertising.02)
```

## 5. The _Credit_ dataset

```{r}
#install.packages("ISLR", repos="https://cloud.r-project.org/")
require(ISLR)
data(Credit)
str(Credit)
```

```{r}
# See the Credit dataset help file 
# From the information we see there, we should:
# 1. Remove the ID from the dataset
# 2. Check the qualitative predictors are indeed coded as factors
Credit<-Credit[,-1]
str(Credit)
```

```{r}
# Isolate the quantitative variables from Credit (as in Figure 3.6)
with(Credit,Credit.Quant<<-data.frame(Balance,Age,Cards,Education, Income,Limit,Rating))
str(Credit.Quant)
```

```{r}
plot(Credit.Quant,pch=19,col="DarkBlue",cex=0.3)
```

```{r}
round(cor(Credit.Quant),2)
```

## 6. Regression with qualitative predictors

### Predicting `Balance` from `Gender`

```{r}
lm.Credit.Gender<-lm(Balance~Gender,data=Credit)
summary(lm.Credit.Gender)
anova(lm.Credit.Gender)
```

```{r}
X1<-model.matrix(lm.Credit.Gender)
```

By default, `lm()` prepares a model matrix for a qualitative predictor entered as a `factor` with $g$ levels, adding $g-1$ columns constructed as follows: it takes one level as the base level and the remaining ones as "treatments" coded with dummy variables, where a $1$ in a given column means that the corresponding treatment is present.

The `contr.*()` functions are internally used to generate these columns. In the above example for the `Credit` data, the two-levels version was used.

```{r}
contr.treatment(2)
contr.treatment(3)
contr.treatment(4)
```

Other codings are possible. For instance, if one does not want to distinguish one level as base:

```{r}
contr.sum(2)
contr.sum(3)
contr.sum(4)
```

```{r}
lm.Credit.Ethnicity<-lm(Balance~Ethnicity,data=Credit)
summary(lm.Credit.Ethnicity)
anova(lm.Credit.Ethnicity)
```

### Interaction terms

```{r}
lm.Advertising.Radio.TV.1<-lm(Sales~Radio+TV,data=Advertising)
summary(lm.Advertising.Radio.TV.1)
anova(lm.Advertising.Radio.TV.1)
```

```{r}
lm.Advertising.Radio.TV.2<-lm(Sales~Radio+TV+Radio:TV,,data=Advertising)
summary(lm.Advertising.Radio.TV.2)
anova(lm.Advertising.Radio.TV.2)
```

```{r}
# Alternative syntax
lm.Advertising.Radio.TV.3<-lm(Sales~Radio*TV,data=Advertising)
summary(lm.Advertising.Radio.TV.3)
anova(lm.Advertising.Radio.TV.3)
```

Interaction of a quantitative predictor and a qualitative predictor in the `Credit` dataset

```{r}
lm.Credit.Income.Student.1<-lm(Balance~Income+Student,data=Credit)
summary(lm.Credit.Income.Student.1)
anova(lm.Credit.Income.Student.1)
```

```{r}
lm.Credit.Income.Student.2<-lm(Balance~Income*Student,data=Credit)
summary(lm.Credit.Income.Student.2)
anova(lm.Credit.Income.Student.2)
```

## 7. The _House prices_ dataset

From Kaggle [House Prices: Advanced Regression Techniques](https://www.kaggle.com/c/house-prices-advanced-regression-techniques/discussion)

```{r}
```